---
title: 'Code to Reproduce Figures and Tables in "A Matched Study Design to Estimate the Effect of Furosemide in Horse Racing"'
author: "Michele Sezgin"
format: pdf
editor: source
---

```{r prep-env, include = T}
library(dplyr)
library(ggplot2)
library(here)
library(brms)
library(tidyr)
library(survival)
library(tidyr)
library(latex2exp)
library(scales)
library(forcats)

set.seed(2025)
theme_set(theme_bw())

source(here('data', 'data-loaders', 'load_data.R'))
```

```{r load-data, include = F}
data_clean <- load_data()
```

## Section 3.1 - Exploratory Data Analysis

### Figure 1

#### a

```{r wwo-plot, echo = T, fig.height = 7, fig.align='center', message = F}
# avg number of with and without
avg_counts <- data_clean |>
  group_by(registration_number, med_description) |>
  summarize(n = n()) |>
  ungroup() |>
  group_by(med_description) |>
  summarize(avg = mean(n) |>
              round(2))

avg_no_lasix <- avg_counts |>
  filter(med_description == 'No Medication') |>
  pull(avg)

avg_lasix <- avg_counts |>
  filter(med_description == 'Lasix') |>
  pull(avg)

data_clean |>
  group_by(registration_number, lasix_ind) |>
  summarize(num_races = n()) |>
  ungroup() |>
  pivot_wider(names_from = lasix_ind, values_from = num_races) |>
  rename(lasix = `1`, no_lasix = `0`) |>
  group_by(lasix, no_lasix) |>
  summarize(count = n(), log_count = log(n())) |>
  ggplot(aes(x = lasix, y = no_lasix, fill = log_count)) +
  geom_tile() +
  labs(x = "Races Without Furosemide",
       y = "Races With Furosemide",
       fill = "Log\nCount") +
  coord_fixed() +
  scale_y_discrete() +
  scale_y_reverse(limits = c(9, 0), breaks = 0:9) +
  scale_x_continuous(position = "top",
                     limits = c(0, 11),
                     breaks = 0:11) +
  scale_fill_continuous(high = "#132B43", low = "#56B1F7") +
  geom_hline(linetype = 'dashed',
             yintercept = avg_lasix,
             color = 'red') +
  geom_vline(linetype = 'dashed',
             xintercept = avg_no_lasix,
             color = 'red') +
  theme(text = element_text(size = 15)) 
```

#### b

```{r odds-plot, message = F, echo = T, fig.height = 4, fig.align='center'}
data_clean |>
  mutate(odds_grouped = cut_number(odds, n = 10, dig.lab = 10)) |>
  group_by(odds_grouped) |>
  summarize(
    avg_itm = mean(in_the_money),
    logit_itm = rje::logit(avg_itm),
    count = n()
  ) |>
  ggplot(aes(x = odds_grouped, y = avg_itm)) +
  geom_point(aes(size = count)) +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1
  )) +
  labs(x = 'Odds', y = 'Proportion ITM', size = 'Count') +
  theme(text = element_text(size = 15)) +
  scale_y_continuous(limits = c(0.2, 0.8))
```

#### c

```{r pp-plot, message = F, echo = T, fig.height = 7, fig.align='center'}
pp_data <- data_clean |>
  mutate(total_horses = total_other_horses + 1) |>
  group_by(post_position, total_horses) |>
  summarize(avg_itm = mean(in_the_money), n = n())

pp_data |>
  ggplot(aes(
    x = post_position,
    y = total_horses,
    color = avg_itm,
    size = n
  )) +
  geom_point(shape = 'square') +
  coord_fixed() +
  scale_color_gradient(high = "#132B43", low = "#56B1F7") +
  labs(x = 'Post Postion',
       y = 'Total Horses',
       size = 'Count',
       color = 'Prop. ITM') +
  theme(text = element_text(size = 15)) +
  scale_y_continuous(limits = c(3, 14), breaks = 3:14) +
  scale_x_continuous(limits = c(1, 14), breaks = 1:14)
```

### d

```{r pitm-plot, fig.width = 6, fig.height = 4, echo = T}
data_clean |>
  mutate(frac_lasix = frac_lasix * .01,
         fl_gr = cut_interval(frac_lasix, n = 10)) |>
  group_by(fl_gr) |>
  summarize(avg_itm = mean(in_the_money), n = n()) |>
  ggplot(aes(x = fl_gr, y = avg_itm, size = n)) +
  geom_point() +
  ylim(0, 1) +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1
  )) +
  labs(x = 'Proportion of Other Horses On Furosemide',
       y = 'Proportion ITM',
       size = 'Count') +
  theme(text = element_text(size = 15))
```

## Section 3.2.1 - Conditional Logistic Regression

### Table 1

```{r load-models, include = T, message = F}
# 3.2.1 conditional logistic regressison
clogit_fit <- readRDS(here('data', 'models', 'clogit_fit.rds'))
# 3.2.2 BHLR
brms_fit_apples <- readRDS(here('data', 'models', 'brms_fit_apples.rds'))
# A.1 BHLR random slopes 
# (not used really, just here for cross-referencing divergent transition +
# variance credible interval nearly containing zero claims)
brms_fit_rs <- readRDS(here('data', 'models', 'brms_fit_rs.rds'))
# A.1 BHLR
brms_fit <- readRDS(here('data', 'models', 'brms_fit.rds'))
```

These are the outputs necessary for reproducing Table 1.

```{r clogit-summary, include = T}
# 3.2.1 conditional logistic
summary(clogit_fit)
exp(confint(clogit_fit))
```

```{r bhlr-apples-summary}
# 3.2.2 BHLR model
summary(brms_fit_apples)
fixef(brms_fit_apples, probs = c(0.025, 0.975)) |>
  exp()
```

## Section 3.2.2 - BHLR

```{r bhlr-preds, include = T}
brms_pred <- readRDS(here('data', 'datasets', 'brms_preds.rds'))
brms_pred_apples <- readRDS(here('data', 'datasets', 'brms_preds_apples.rds'))
```

```{r cal-data, message = F, warning = F, echo = T}
calibration_data <- data.frame(
  clogit = predict(clogit_fit, type = 'lp', reference = 'strata') |>
    rje::expit(),
  brms = brms_pred[, 1],
  brms_apples = brms_pred_apples[, 1],
  actual = data_clean$in_the_money
) |>
  tidyr::pivot_longer(c(clogit, brms, brms_apples),
                      names_to = 'model',
                      values_to = 'pred') |>
  mutate(model = case_when(
    model == 'clogit' ~ 'Cond. Logit',
    model == 'brms_apples' ~ 'BHLR',
    # this is the BHLR fit with extra random effects
    # terms from A.1
    T ~ 'Model 3'
  ))
```

### Figure 2

```{r cal-plot-3-2, message = F, echo = T, fig.height = 4, fig.width = 6}
#| fig.cap: 'Predicted probabilities binned into 0.1-wide intervals.'
calibration_data |>
  filter(model == 'BHLR') |>
  mutate(interval = cut(
    pred,
    breaks = seq(0, 1, by = .1),
    include.lowest = T
  )) |>
  group_by(interval) |>
  summarize(
    int_pred = mean(pred),
    int_actual = mean(actual),
    se = sqrt((int_actual * (1 - int_actual)) / n()),
    n = n(),
    ci_lower = pmax(int_actual - 2 * se, 0),
    ci_upper = pmin(int_actual + 2 * se, 1)
  ) |>
  ggplot(aes(x = int_pred, y = int_actual)) +
  geom_point(alpha = 0.5, aes(size = n)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper)) +
  geom_smooth(se = FALSE,
              method = 'loess',
              color = 'black') +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dashed") +
  labs(x = "Predicted probability",
       y = "Observed fraction",
       color = 'Model',
       size = 'Count') +
  ylim(0, 1) +
  theme(text = element_text(size = 15))
```

## Section A.1.1 Results

This is not used to produce any figures or tables but shows the results of
fitting the random slopes model described in section A.1:

```{r bhlr-rs-summary}
# BHLR random slopes model described in sec A.1 
summary(brms_fit_rs)
```

### Figure A.1

```{r bhlr-data-trace, include = T, message = F}
# extract the data necessary to make trace plots: samples and iterations.
brms_apples_data <- brms_fit_apples$fit@sim$samples |>
  purrr::list_rbind(names_to = 'index') |>
  group_by(index) |>
  mutate(iter = 1:n()) |>
  ungroup()

brms_data <- brms_fit$fit@sim$samples |>
  purrr::list_rbind(names_to = 'index') |>
  group_by(index) |>
  mutate(iter = 1:n()) |>
  ungroup()
```

```{r forest-plot-data-apples, message = F, echo = T, fig.height = 4, fig.width = 6}
# forest plots
lower <- brms_apples_data |>
  dplyr::select(
    'b_med_descriptionLasix',
    'b_frac_lasix',
    'b_total_other_horses',
    'b_post_position'
  ) |>
  sapply(quantile, probs = .025) |>
  sapply(exp) |>
  unname()

upper <- brms_apples_data |>
  dplyr::select(
    'b_med_descriptionLasix',
    'b_frac_lasix',
    'b_total_other_horses',
    'b_post_position'
  ) |>
  sapply(quantile, probs = .975) |>
  sapply(exp) |>
  unname()

means <- brms_apples_data |>
  dplyr::select(
    'b_med_descriptionLasix',
    'b_frac_lasix',
    'b_total_other_horses',
    'b_post_position'
  ) |>
  colMeans() |>
  sapply(exp) |>
  unname()

# names <- brms_apples_data |>
#   dplyr::select(starts_with('b')) |>
#   colnames()

names <- fct_inorder(c(
  'Furosemide',
  'Frac. Furosemide',
  'Total Other Horses',
  'Post Position'
))

brms_apples_forest_data <- tibble(
  coef = names,
  mean = means,
  lower = lower,
  upper = upper
)
```

```{r forest-plot-bhlr, message = F, echo = T, fig.height = 3, fig.width = 6}
lower <- brms_data |>
  dplyr::select('b_med_descriptionLasix',
                'b_frac_lasix',
                'b_total_other_horses') |>
  sapply(quantile, probs = .025) |>
  sapply(exp) |>
  unname()

upper <- brms_data |>
  dplyr::select('b_med_descriptionLasix',
                'b_frac_lasix',
                'b_total_other_horses') |>
  sapply(quantile, probs = .975) |>
  sapply(exp) |>
  unname()

means <- brms_data |>
  dplyr::select('b_med_descriptionLasix',
                'b_frac_lasix',
                'b_total_other_horses') |>
  colMeans() |>
  sapply(exp) |>
  unname()

names <- fct_inorder(c('Furosemide',
                       'Frac. Furosemide',
                       'Total Other Horses'))

brms_forest_data <- tibble(
  coef = names,
  mean = means,
  lower = lower,
  upper = upper,
  model = 'Model 3'
)

combined_data <- brms_apples_forest_data |>
  mutate(model = 'Model 2') |>
  bind_rows(brms_forest_data)

ggplot(data = combined_data) +
  geom_vline(
    xintercept = 1,
    linetype = 'dashed',
    color = 'red',
    linewidth = .75,
    alpha = .5
  ) +
  geom_errorbar(
    aes(
      xmin = lower,
      xmax = upper,
      y = fct_rev(coef),
      color = model
    ),
    linewidth = .75,
    alpha = .5
  ) +
  geom_point(
    shape = 21,
    fill = 'white',
    aes(x = mean, y = fct_rev(coef), color = model),
    size = 3,
    alpha = 0.5
  ) +
  labs(x = 'Odds Ratio', y = '', color = 'Model')
```

### Figure A.2

```{r bhlr-2-3-cal-plot, message = F, echo = T, fig.height = 4, fig.width = 6}
# #| fig.cap: 'Predicted probabilities binned into 0.1-wide intervals.'
calibration_data |>
  filter(model != 'Cond. Logit') |>
  mutate(
    model = if_else(model == 'BHLR', 'Model 2', model),
    interval = cut(
      pred,
      breaks = seq(0, 1, by = .1),
      include.lowest = T
    )
  ) |>
  group_by(interval, model) |>
  summarize(
    int_pred = mean(pred),
    int_actual = mean(actual),
    se = sqrt((int_actual * (1 - int_actual)) / n()),
    n = n(),
    ci_lower = pmax(int_actual - 2 * se, 0),
    ci_upper = pmin(int_actual + 2 * se, 1)
  ) |>
  ggplot(aes(x = int_pred, y = int_actual, color = model)) +
  geom_point(alpha = 0.5, aes(size = n)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper)) +
  geom_smooth(se = FALSE, method = 'loess') +
  geom_abline(slope = 1,
              intercept = 0,
              linetype = "dashed") +
  labs(x = "Predicted probability",
       y = "Observed fraction",
       color = 'Model',
       size = 'Count') +
  ylim(0, 1) +
  theme(text = element_text(size = 15))
```

### Table A.1

```{r bhlr-summary}
# BHLR model described in sec A.1 and eq (A.1)
summary(brms_fit)
```

## Section A.1.2 Randomized Quantile Residuals

### Figure A.3 

#### a

```{r rqr-bhlr-apples, echo = T, message = F, fig.height = 5, fig.width = 6}
rqr_tbl <- data_clean |>
  mutate(brms_pred_apples = brms_pred_apples,
         cdf_bound = 1 - brms_pred_apples) |>
  rowwise() |>
  mutate(rq_resid = if_else(
    in_the_money == 0,
    runif(1, min = 0, max = cdf_bound),
    runif(1, min = cdf_bound, max = 1)
  )) |>
  ungroup()

ggplot(rqr_tbl, aes(x = odds, y = rq_resid)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  geom_hline(linetype = 'dashed',
             color = 'red',
             yintercept = .5) +
  ylim(0, 1) +
  theme(text = element_text(size = 15)) +
  labs(x = 'Odds', y = 'Rand. Quant. Residuals')
```

#### b

```{r rqr-bhlr, echo = T, message = F, fig.height = 5, fig.width = 6}
rqr_tbl <- data_clean |>
  mutate(brms_pred = brms_pred, cdf_bound = 1 - brms_pred) |>
  rowwise() |>
  mutate(rq_resid = if_else(
    in_the_money == 0,
    runif(1, min = 0, max = cdf_bound),
    runif(1, min = cdf_bound, max = 1)
  )) |>
  ungroup()

ggplot(rqr_tbl, aes(x = odds, y = rq_resid)) +
  geom_point(alpha = 0.2) +
  geom_smooth(se = FALSE) +
  geom_hline(linetype = 'dashed',
             color = 'red',
             yintercept = .5) +
  ylim(0, 1) +
  theme(text = element_text(size = 15)) +
  labs(x = 'Odds', y = 'Rand. Quant. Residuals')
```

\vspace{200pt}

## Section A.1.3 BHLR Trace Plots

### Figure A.4

#### a

```{r bhlr-apples-trace-var, fig.height = 5, fig.width = 6, echo = T}
ggplot(
  brms_apples_data |>
    mutate(index = paste0('Chain ', index)),
  aes(x = iter, y = sd_registration_number__Intercept)
) +
  geom_line() +
  facet_grid(rows = vars(index)) +
  labs(y = TeX('$\\hat{\\sigma}_\\alpha$'),
       x = 'Iteration')
```

#### b

```{r bhlr-apples-trace-md, fig.height = 5, fig.width = 6, echo = T}
ggplot(brms_apples_data |>
         mutate(index = paste0('Chain ', index)),
       aes(x = iter, y = b_med_descriptionLasix)) +
  geom_line() +
  facet_grid(rows = vars(index)) +
  labs(y = TeX('$\\hat{\\tau}$'),
       x = 'Iteration')
```

### Figure A.5

#### a

```{r bhlr-trace-var, fig.height = 5, fig.width = 6, echo = T}
ggplot(
  brms_data |>
    mutate(index = paste0('Chain ', index)),
  aes(x = iter, y = sd_registration_number__Intercept)
) +
  geom_line() +
  facet_grid(rows = vars(index)) +
  labs(y = TeX('$\\hat{\\sigma}_\\alpha$'),
       x = 'Iteration')
```

#### b

```{r bhlr-trace-md, fig.height = 5, fig.width = 6, echo = T}
ggplot(brms_data |>
         mutate(index = paste0('Chain ', index)),
       aes(x = iter, y = b_med_descriptionLasix)) +
  geom_line() +
  facet_grid(rows = vars(index)) +
  labs(y = TeX('$\\hat{\\tau}$'),
       x = 'Iteration')
```

...